---
title: "Kirc High Variance"
author: "Kevin Feng"
date: '2022-06-10'
output:
  html_document:
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preparation
### Load Data, Calculate Variance
This section only needs to run once. Run just the next section for future work so we do not need to recalculate variance as that takes a lot of time and memory.
```{r dataPrep, message=FALSE, eval=FALSE}
library(matrixStats)
library(data.table)

data = readRDS('kirc_824.RDS') # [1] 4022634     824
class(data) # this is matrix when we read it in 

# get the following for each feature
variance <- rowVars(data) # takes a while and needs ~25gb of ram
rowIndex <- c(1:nrow(data))

# compare variance here to python variance
data.var.filtered <- fread('kirc_824_varFiltered_cutoff_4.csv')
dim(data.var.filtered) # [1] 9693  826, 2 extra columns for featureNames and var
data.var.filtered[data.var.filtered$featureNames == '15939']$featureVar == variance[15939]

max(variance) # this looks the same as the python output
sum(variance >= 4) # this also gives the same number as python

# test on small subset to see if constructing a df works
test <- as.data.frame(data[1:5, 1:5])
test$idx <- rowIndex[1:5]
test$var <- variance[1:5]
test # works!

# construct whole df
data.df <- as.data.frame(data)
data.df$idx <- rowIndex
data.df$var <- variance
dim(data.df) # [1] 4022634     826
data.df[1:5, 820:826]

# load sample info
sample.info = readRDS('sample_info.RDS')

# write to RDS
saveRDS(data.df, file = 'kirc824_variance_idx.RDS')
```


### Save Compute Time
```{r saveTime, message=FALSE}
library(matrixStats)
library(data.table)
data <- readRDS('kirc824_variance_idx.RDS')
data.df <- as.data.frame(data)

# load sample info
sample.info = readRDS('sample_info.RDS')
```


### Subset for High Variance Features
```{r high.var.subset, message=FALSE}
library(dplyr)
data.df.filtered <- data.df %>% filter(var >= 4) 
dim(data.df.filtered) # [1] 9693  826
data.df.filtered[1:5, 820:826] # last 2 columns should not be used in pca

# set rownames of the df to idx
rownames(data.df.filtered) <- data.df.filtered$idx
```


## Analysis
### Boxplots with Scatters Function and Data Prep
```{r boxplots.w.scatters, message=FALSE}
library(hrbrthemes)
library(viridis)
library(gridExtra)
library(grid)
library(ggpubr)
library(shiny)
library(ggplot2)
library(tidyverse)
library(factoextra)

# plots boxplots with scatters with given data and list of features of interest
# data must be in form of df samples columns + idx column, features in rows
# class.lables = class labels for samples
# pc.name to title plot
plot.boxplots.scatters <- function(data, feature.list, class.labels, plot.name, n.row=2){
  # subset for only data features of interest
  data <- data %>% filter(idx %in% as.character(feature.list))
  # set row names, drop idx
  rownames(data) <- data$idx
  data <- data %>% select(-idx)
  
  # transpose, must cast as dataframe to add class
  data <- as.data.frame(t(data))
  
  # add class labels as column
  data$class <- class.labels
  
  # melt and plot
  data.m <- melt(data, id.vars = 'class')
  ggplot(data.m, aes(class, value, fill=class), alpha=0.5) + geom_boxplot(outlier.shape = NA) + scale_fill_brewer(palette="Pastel2") + theme_minimal() + theme(axis.text.x=element_blank()) +
      geom_jitter(color="black", size=0.3, alpha=0.5) + facet_wrap(~variable, nrow=n.row) + ggtitle(plot.name)
}



# drop var column
data.df.filtered.no.var <- data.df.filtered %>% select(-var)
```


## Normal Vs Tumor Comparisons
### Split the Data
```{r dataSplit, message=FALSE}
# we need to start with the filtered dataset, append sample info, then select
# remove var and idx
data.df.filtered.new <- select(data.df.filtered, -c(idx,var)) 

# set class info to a var called class
class <- sample.info[,2]

# subset for normal samples
normal.df <- data.df.filtered.new[, class == 'normal']

# subset for tumor samples
tumor.df <- data.df.filtered.new[, class == 'tumor']

# check to make sure they're the right size
dim(normal.df) # [1] 9693  289, loses idx so we need to add it back in 
sum(class == 'normal') # 289

# repeat for tumor
dim(tumor.df) # [1] 9693  536, contains idx
sum(class == 'tumor') # [1] 535


# check a random sample from each
all(normal.df$subj_111_normal == data.df.filtered$subj_111_normal)
all(tumor.df$subj_111_tumor == data.df.filtered$subj_111_tumor)

# check that idx's aka feature names are the same for both dataframes
all(rownames(normal.df) == rownames(tumor.df))
```


### Calculate Quantiles
```{r quantileCalc, message=FALSE}
# calc normal tissue quantiles
normal.q5 <- rowQuantiles(as.matrix(normal.df), probs = 0.05)
normal.q50 <- rowQuantiles(as.matrix(normal.df), probs = 0.50)
normal.q95 <- rowQuantiles(as.matrix(normal.df), probs = 0.95)

# calc tumor quantiles
tumor.q5 <- rowQuantiles(as.matrix(tumor.df), probs = 0.05)
tumor.q50 <- rowQuantiles(as.matrix(tumor.df), probs = 0.50)
tumor.q95 <- rowQuantiles(as.matrix(tumor.df), probs = 0.95)

# instances where the tumor q5 is greater than the normal q95
sum(tumor.q5>normal.q95)

## To do: subset both df's for these features and plot
```


### Subset the split DFs
```{r subsetSplit, message=FALSE}
# append quantiles as columns to each df
# normal
normal.df$q5 <- normal.q5
normal.df$q50 <- normal.q50
normal.df$q95 <- normal.q95

# tumor
tumor.df$q5 <- tumor.q5
tumor.df$q50 <- tumor.q50
tumor.df$q95 <- tumor.q95

# add idx back in since the rows have not been messed with they'll stay in order
# we need this because dplyr::filter does not retain rownames
normal.df$idx <- data.df.filtered$idx
tumor.df$idx <- data.df.filtered$idx

tumor.q5.normal.q95 <- tumor.df %>% filter(q5 > normal.df$q95)

# save list of features of interest as a variable
big.spread.features <- tumor.q5.normal.q95$idx
```


### Boxplots with Scatters T q5 > N q95
```{r tumorq5.normalq95.boxplots.w.scatters, message=FALSE, fig.height=7}
# below: previous code to plot, replaced with function in above section
# # subset high var df for tumor q5 > q95 features
# subset.tq5.nq95 <- data.df.filtered %>% filter(idx %in% as.character(big.spread.features))
# 
# # drop the var column
# subset.tq5.nq95 <- subset.tq5.nq95 %>% select(-var)
# 
# # set row names, drop idx
# rownames(subset.tq5.nq95) <- subset.tq5.nq95$idx
# subset.tq5.nq95 <- subset.tq5.nq95 %>% select(-idx)
# 
# # transpose, must cast as dataframe to add class
# subset.tq5.nq95.t <- as.data.frame(t(subset.tq5.nq95))
# 
# # add class labels 
# subset.tq5.nq95.t$class <- class
# 
# # melt
# subset.tq5.nq95.t.m <- melt(subset.tq5.nq95.t, id.vars = 'class')
# 
# # plot
# ggplot(subset.tq5.nq95.t.m, aes(class, value, fill=class)) + geom_boxplot() + scale_fill_viridis(discrete = TRUE, alpha=0.6) +
#       geom_jitter(color="black", size=0.2, alpha=0.9) + facet_wrap(~variable, nrow=2)

plot.boxplots.scatters(data.df.filtered.no.var, big.spread.features, sample.info[,2], 'Features Tumor q5 > Normal q95', 4)
```


### Check Different Cutoffs
```{r diffCutoffs, message=FALSE}
# vary cutoffs and see how many features we get
# need to exclude the last 4 columns in both dfs are the are q5, q50, q95, and idx
for (q in seq(0.01, 0.1, 0.01)){
  normal.q <- rowQuantiles(as.matrix(normal.df[, 1:(ncol(normal.df)-4)]), probs = 1-q)

  tumor.q <- rowQuantiles(as.matrix(tumor.df[, 1:(ncol(tumor.df)-4)]), probs = 0+q)
  
  cat('Tumor q', 0+q, '> Normal q', 1-q, ':', sum(tumor.q>normal.q), '\n')
}
  
```

### Tumor q6 > Normal q94 Box Plot 
```{r tumorq6.normalq94, message=FALSE, fig.height=8}
# calc tumor q6
normal.q94 <- rowQuantiles(as.matrix(normal.df[, 1:(ncol(normal.df)-4)]), probs = 0.94)

tumor.q6 <- rowQuantiles(as.matrix(tumor.df[, 1:(ncol(tumor.df)-4)]), probs = 0.06)

tumor.q6.normal.q94 <- tumor.df %>% filter(tumor.q6 > normal.q94)

# save list of features of interest as a variable
q6.q94.features <- tumor.q6.normal.q94$idx

# plot
plot.boxplots.scatters(data.df.filtered.no.var, q6.q94.features, sample.info[,2], 'Features Tumor q4 > Normal q94', 7)
```


### Tumor q10 > Normal q90 Box Plot
```{r tumorq10.normalq90, message=FALSE, fig.height=15}
# calc tumor q6
normal.q90 <- rowQuantiles(as.matrix(normal.df[, 1:(ncol(normal.df)-4)]), probs = 0.90)

tumor.q10 <- rowQuantiles(as.matrix(tumor.df[, 1:(ncol(tumor.df)-4)]), probs = 0.10)

tumor.q10.normal.q90 <- tumor.df %>% filter(tumor.q10 > normal.q90)

# save list of features of interest as a variable
q10.q90.features <- tumor.q10.normal.q90$idx

# plot
plot.boxplots.scatters(data.df.filtered.no.var, q10.q90.features, sample.info[,2], 'Features Tumor q10 > Normal q90', 14)
```


## Further Filtering 
### Variance difference > 1

We will start with the 129 t q10 > n q90 features and try to filter that down to 30-40.
```{r varDiff1, message=FALSE, fig.height=10}
# subset whole data for the 129 features 
data.df.tq10.nq90 <- data.df.filtered.no.var %>% filter(idx %in% q10.q90.features)

# mean difference between feature expression for the 2 groups
mean(tumor.q10[names(tumor.q10) %in% q10.q90.features] - normal.q90[names(normal.q90) %in% q10.q90.features])

diff.var.1.features <- q10.q90.features[(tumor.q10[names(tumor.q10) %in% q10.q90.features] - normal.q90[names(normal.q90) %in% q10.q90.features]) > 1]

plot.boxplots.scatters(data.df.filtered.no.var, diff.var.1.features, sample.info[,2], 'Var diff > 1', 6)
```


### Tumor 'min' > Normal 'max'
```{r iqr.min.max, message=FALSE, fig.height=10}
# we want tumor 'min' = Q1(25th) - 1.5*IQR > normal 'max' = Q5(75th) + 1.5*IQR
# subset tumor and normal df's for the 129 features so we can compute
tumor.df.129 <- tumor.df %>% filter(idx %in% q10.q90.features)
normal.df.129 <- normal.df %>% filter(idx %in% q10.q90.features)

# compute iqr for each row
tumor.df.129$iqr <- rowIQRs(as.matrix(tumor.df.129[,1:(ncol(tumor.df.129)-4)]))
normal.df.129$iqr <- rowIQRs(as.matrix(normal.df.129[,1:(ncol(normal.df.129)-4)]))

# we can play around with the quantile used for each
# calc min for tumor 
# can lower probs to make this more stringent 
tumor.df.129$min <- rowQuantiles(as.matrix(tumor.df.129[, 1:(ncol(tumor.df.129)-5)]), probs = 0.25) - (1.5*tumor.df.129$iqr) 

# calc max for normal
# can raise probs to make this more stringent 
normal.df.129$max <- rowQuantiles(as.matrix(normal.df.129[, 1:(ncol(normal.df.129)-5)]), probs = 0.75) + (1.5*normal.df.129$iqr) 

# how many features do we have that meet the criteria
sum(tumor.df.129$min > normal.df.129$max) # 22

plot.boxplots.scatters(data.df.filtered.no.var, tumor.df.129[tumor.df.129$min > normal.df.129$max,]$idx, sample.info[,2], 
                       'T q10 > N q90 + T \'min\' > N \'max\'', 3)
```


### Min/max All High Variance Features (9693)
```{r moreFiltering2, message=FALSE, fig.height=12}
tumor.df.iqr <- tumor.df
normal.df.iqr <- normal.df

# compute iqr for each row
tumor.df.iqr$iqr <- rowIQRs(as.matrix(tumor.df[,1:(ncol(tumor.df)-4)]))
normal.df.iqr$iqr <- rowIQRs(as.matrix(normal.df[,1:(ncol(normal.df)-4)]))

# we can play around with the quantile used for each
# calc min for tumor 
# can lower probs to make this more stringent 
tumor.df.iqr$min <- rowQuantiles(as.matrix(tumor.df.iqr[, 1:(ncol(tumor.df.iqr)-5)]), probs = 0.25) - (1.5*tumor.df.iqr$iqr) 

# calc max for normal
# can raise probs to make this more stringent 
normal.df.iqr$max <- rowQuantiles(as.matrix(normal.df.iqr[, 1:(ncol(normal.df.iqr)-5)]), probs = 0.75) + (1.5*normal.df.iqr$iqr) 

# how many features do we have that meet the criteria
sum(tumor.df.iqr$min > normal.df.iqr$max) # 22

plot.boxplots.scatters(data.df.filtered.no.var, tumor.df.iqr[tumor.df.iqr$min > normal.df.iqr$max,]$idx, sample.info[,2], 
                       'T \'min\' > N \'max\'', 4)
```

In the above plot there are more outliers that dip into the distribution of the other class. I believe this t min > n max works better in conjunction with t q10 > n q90, than by itself. 

We can try having an outlier limit of 10%.

### Outlier Limit
```{r outlierCutoff, message=FALSE, fig.height=10}

# tumor.df.iqr['2430535', (ncol(tumor.df.iqr)-5):ncol(tumor.df.iqr)]

# outlier = value < iqr min 
# count number of outliers for each feature
n.outliers.tumor <- list()
for (i in 1:nrow(tumor.df.iqr)){
  this.outliers <- sum(tumor.df.iqr[i, 1:(ncol(tumor.df.iqr)-6)] < tumor.df.iqr$min)
  n.outliers.tumor <- c(n.outliers.tumor, this.outliers)
}

# outlier = value > iqr max
# count number of outliers for each feature
n.outliers.normal <- list()
for (i in 1:nrow(normal.df.iqr)){
  this.outliers <- sum(normal.df.iqr[i, 1:(ncol(normal.df.iqr)-6)] > normal.df.iqr$max)
  n.outliers.normal <- c(n.outliers.normal, this.outliers)
}

# add to data frame
tumor.df.iqr$n.outliers <- unlist(n.outliers.tumor, recursive=FALSE)

normal.df.iqr$n.outliers <- unlist(n.outliers.normal, recursive=FALSE)

# check to make sure everything looks right, we can use Feature 2430535 since we already counted those outliers
tumor.df.iqr[1:5, (ncol(tumor.df.iqr)-5):ncol(tumor.df.iqr)]
tumor.df.iqr['2430535', (ncol(tumor.df.iqr)-5):ncol(tumor.df.iqr)]


# set cutoffs, lets use 50 for tumors and 20 for normal (approx. 10%)
tumor.df.outlier.filter <- tumor.df.iqr %>% filter(n.outliers <= 50)

normal.df.outlier.filter <- normal.df.iqr %>% filter(n.outliers <= 20)

# now we need to get overlapping idx's from each df
outlier.filter.feature.list <- intersect(tumor.df.outlier.filter$idx, normal.df.outlier.filter$idx) 
length(outlier.filter.feature.list) # 120 features that meet this condition

# lets see how many also meet t q10 > n q90
intersect(outlier.filter.feature.list, q10.q90.features)

# try plotting these
plot.boxplots.scatters(data.df.filtered.no.var, intersect(outlier.filter.feature.list, q10.q90.features), sample.info[,2],
                       't q10 > n q90 + low outliers', 4)
```

### Limit outliers with stronger cutoffs
```{r outliers.q25.q75, message=FALSE, fig.height=10}
# read in the tumor and normal df's
tumor.df.q25 <- tumor.df
normal.df.q75 <- normal.df

# calculate respective quantiles for each df 
tumor.df.q25$q25 <- rowQuantiles(as.matrix(tumor.df.q25), probs = 0.25)
normal.df.q75$q75 <- rowQuantiles(as.matrix(normal.df.q75), probs = 0.75)

# outlier = value < q25
# count number of outliers for each feature
n.outliers.tumor <- list()
for (i in 1:nrow(tumor.df.q25)){
  this.outliers <- sum(tumor.df.q25[i, 1:(ncol(tumor.df.q25)-5)] <= tumor.df.q25$q25)
  n.outliers.tumor <- c(n.outliers.tumor, this.outliers)
}

# outlier = value > q75
# count number of outliers for each feature
n.outliers.normal <- list()
for (i in 1:nrow(normal.df.q75)){
  this.outliers <- sum(normal.df.q75[i, 1:(ncol(normal.df.q75)-5)] >= normal.df.q75$q75)
  n.outliers.normal <- c(n.outliers.normal, this.outliers)
}

# add to data frame
tumor.df.q25$n.outliers <- unlist(n.outliers.tumor, recursive=FALSE)

normal.df.q75$n.outliers <- unlist(n.outliers.normal, recursive=FALSE)

# set cutoffs, lets use 60 for tumors and 30 for normal (approx. 10%)
tumor.df.outlier.filter <- tumor.df.q25 %>% filter(n.outliers <= 75) # 535 tumor samples

normal.df.outlier.filter <- normal.df.q75 %>% filter(n.outliers <= 60) # 289 normal samples

# now we need to get overlapping idx's from each df
outlier.filter.feature.list <- intersect(tumor.df.outlier.filter$idx, normal.df.outlier.filter$idx) 
length(outlier.filter.feature.list) # 10 features that meet this condition

# plot feature that meet the condition 
plot.boxplots.scatters(data.df.filtered.no.var, outlier.filter.feature.list, sample.info[,2],
                       't q25, n q75 outlier limits', 1)


# number of features that meet the above condition and have t iqr min > n iqr max
length(intersect(outlier.filter.feature.list, tumor.df.iqr[tumor.df.iqr$min > normal.df.iqr$max,]$idx))

# plot features that meet both conditions 
plot.boxplots.scatters(data.df.filtered.no.var, 
                       intersect(outlier.filter.feature.list, tumor.df.iqr[tumor.df.iqr$min > normal.df.iqr$max,]$idx), 
                       sample.info[,2],
                       't q25, n q75 outlier limits + t iqr min > n iqr max', n.row = 1)
```

### Original Dataset var > 3 cutoff
```{r anotherTest, message=FALSE, fig.height=10}
# subset original data with a lower variance threshold
data.df.var3 <- data.df %>% filter(var > 3)

# subset for normal samples
normal.df.var3 <- data.df.var3[, class == 'normal']

# subset for tumor samples
tumor.df.var3 <- data.df.var3[, class == 'tumor']

# t min iqr > n max iqr
# compute iqr for each row
tumor.df.var3$iqr <- rowIQRs(as.matrix(tumor.df.var3[,1:(ncol(tumor.df.var3)-2)]))
normal.df.var3$iqr <- rowIQRs(as.matrix(normal.df.var3[,1:(ncol(normal.df.var3)-2)]))

# we can play around with the quantile used for each
# calc min for tumor 
# can lower probs to make this more stringent 
tumor.df.var3$min <- rowQuantiles(as.matrix(tumor.df.var3[, 1:(ncol(tumor.df.var3)-3)]), probs = 0.25) - (1.5*tumor.df.var3$iqr) 

# calc max for normal
# can raise probs to make this more stringent 
normal.df.var3$max <- rowQuantiles(as.matrix(normal.df.var3[, 1:(ncol(normal.df.var3)-3)]), probs = 0.75) + (1.5*normal.df.var3$iqr) 

sum(tumor.df.var3$min > normal.df.var3$max) # 38 features

# save idx that meet pass the iqr min max cutoff 
iqr.cutoff.idx <- tumor.df.var3$idx[tumor.df.var3$min > normal.df.var3$max] 

# drop var column for boxplots
data.df.var3.to.plot <- data.df.var3%>% select(-var)

plot.boxplots.scatters(data.df.var3.to.plot, iqr.cutoff.idx, class, 'Var > 3, T iqr min > N iqr max', n.row = 4)

# compute outliers 
out <- sapply(tumor.df.var3, function(x) x <= quantile(x, 0.25))
tumor.df.var3$outlier_count <- rowSums(out)

out <- sapply(normal.df.var3, function(x) x >= quantile(x, 0.75))
normal.df.var3$outlier_count <- rowSums(out)

# make sure all idx are still order
all(tumor.df.var3$idx == normal.df.var3$idx)

# sum of outliers, we only care about lows in tumors and highs in norms 
sum.of.outliers = normal.df.var3$outlier_count + tumor.df.var3$outlier_count

# set var for idx
idx <- tumor.df.var3$idx

# makea df with idx and sum of outlierss
idx.outlier.df <- as.data.frame(cbind(idx, sum.of.outliers))

# sort by sum of outliers
idx.outlier.df.sorted <- idx.outlier.df[order(sum.of.outliers),]

intersect(idx.outlier.df.sorted$idx, iqr.cutoff.idx) # intersect returns in order of the first input

# filter for the 38 idx's of interest 
idx.outliers.filtered <- idx.outlier.df.sorted %>% filter(idx %in% iqr.cutoff.idx)
idx.outliers.filtered
# drop the last 8 to get 30 features
head(idx.outliers.filtered$idx, -6)

plot.boxplots.scatters(data.df.var3.to.plot, head(idx.outliers.filtered$idx, -8), class, 'Var > 3, T iqr min > N iqr max, low outliers', n.row = 4)
```


## Check Feature for Outliers
### Feature 2430535
```{r checkFeature.2430535}
# checking outliers for feature 2430535
test <- (data.df.filtered.no.var['2430535',1:(ncol(data.df.filtered.no.var)-1)])

# split for class
t.t <- test[, (sample.info[,2]=='tumor')]
t.n <- test[, (sample.info[,2]=='normal')]

# plot distributions 
hist(t(t.t), main = 'Tumor Distribution of Feature 2430535', xlab = 'Expression', breaks = 20)

hist(t(t.n), main = 'Normal Distribution of Feature 2430535', xlab = 'Expression', breaks = 20)

# check how many tumor samples above 2
sum(t.t > 2) # 124
dim(t.t) # 535

# ~124 of the 535 samples could be considered outliers 
```



### Make Text File 
Contains feature name, T qX and N qY
```{r makeFile, message=FALSE, eval=FALSE}
txtfile <- as.data.frame(cbind(data.df.filtered$idx, tumor.q10, normal.q90))
txtfile.filt <- txtfile[txtfile$tumor.q10>txtfile$normal.q90,]
colnames(txtfile.filt)[1] <- 'featureIdx'
all(txtfile.filt$featureIdx == (tumor.q10.normal.q90$idx))

write.table(txtfile.filt, file = 'tumorq10_greaterThan_normalq90.txt', sep = ' ', row.names = FALSE)
```




